``Interestingness'' in Mathematical Theory Exploration

### 2019-10-17 PhDThesis/litreview_staticanalysis.tex
More static analysis

A more speculative use of optimisation during runtime has been proposed by
Hutter~\iffalse TODO: Cite \fi. As in a JIT-enabled interpreter, Hutter's
algorithm begins by directly executing the given (unoptimised) program $p$ on
the relevant input $x$ (this may be provided up front, or piecemeal throughout
execution, e.g. as a stream of user interaction events). The variables $p_{fast}$
and $t_{fast}$ are initialised to $p$ and $\infinity$ respectively, to indicate
that $p$ is the fastest implementation we have so far, and we don't yet have an
upper bound on its running time. Concurrently with the execution of $p$, two
other processes are also started:

@@

Hutter's algorithm is mostly of theoretical interest, being described as
``another $O()$ warning showing, how important [to running time] factors, and
even subdominant additive terms, are.''~\cite{TODO: Hutter, page 4}. Despite
this caveat, the general framework is still useful to compare with the code
improvement techniques described above. In particular, the use of open-ended
search (process A) is similar to that of superoptimisation; although the former
enumerates equivalence proofs and extracts programs from them, whilst the latter
enumerates programs and attempts to prove (or approximately determine)
equivalence. Likewise the speed estimation of process B is analogous to those
employed by superoptimisation. JIT compilation can also be achieved within this
framework, for example by including our desired optimisations in the axioms of
the formal system (making proof search trivial), biasing the search algorithm to
consider applications of these optimisations first, and including a simple cost
model for the language. Switching to faster programs can also be made
incremental, e.g. allowing (stateless) functions to be replaced individually
without having to restart execution, just like in a JIT-enabled interpreter.

The open-ended search through program and proof space performed by
superoptimisation and Hutter's algorithm are similar to that of Theory
Exploration (e.g. the discovery of equivalent expressions by \isacosy{} and
\quickspec{}). One major difference is in \emph{end result} of these techniques:
the latter tends to \emph{build up} structure and abstraction latent in a set of
definitions, whilst the former tend to \emph{tear down} such layers to find
minimal implementations.

Another seeming contrast is that Theory Exploration has no ``target'', unlike
optimisers. However, the practical necessity of limiting superoptimisers to only
small snippets of code at a time has resulted in applications like finding
``peephole optimisations''~\cite{Bansal.Aiken:2006}, whose scattershot nature is
more akin to Theory Exploration than traditional compiler optimisation.

In any case, efficiency (whether instruction count, small upper bounds, cost
semantics, etc.) can be considered part of the interestingness criteria of
Theory Exploration, and there are certainly opportunities to cross-pollinate
techniques between these fields.

% TODO: Evolutionary patches
% TODO: Goedel machines

>>>>

A more speculative use of optimisation during runtime has been proposed by
Hutter~\iffalse TODO: Cite \fi. As in a JIT-enabled interpreter, Hutter's
algorithm (also referred \hsearch{}) begins by directly executing the given
(unoptimised) program $p$ on the relevant input $x$ (this may be provided up
front, or piecemeal throughout execution, e.g. as a stream of user interaction
events). The variables $p_{fast}$ and $t_{fast}$ are initialised to $p$ and
$\infinity$ respectively, to indicate that $p$ is the fastest implementation we
have so far, and we don't yet have an upper bound on its running
time. Concurrently with the execution of $p$, two other processes are also
started:

@@

\hsearch{} is mostly of theoretical interest, being described as ``another $O()$
warning showing, how important [to running time] factors, and even subdominant
additive terms, are.''~\cite{TODO: Hutter, page 4}. Despite this caveat, the
general framework is still useful to compare with the code improvement
techniques described above. In particular, the use of open-ended search (process
A) is similar to that of superoptimisation; although the former enumerates
equivalence proofs and extracts programs from them, whilst the latter enumerates
programs and attempts to prove (or approximately determine)
equivalence. Likewise the speed estimation of process B is analogous to those
employed by superoptimisation. JIT compilation can also be achieved within this
framework, for example by including our desired optimisations in the axioms of
the formal system (making proof search trivial), biasing the search algorithm to
consider applications of these optimisations first, and including a simple cost
model for the language. Switching to faster programs can also be made
incremental, e.g. allowing (stateless) functions to be replaced individually
without having to restart execution, just like in a JIT-enabled interpreter.

The framework of the \goedelmachine{}~\cite{Schmidhuber:05icann} goes even
further than \hsearch{}: rather than limiting optimisations to just the given
program, it also searches for provable optimisations to \emph{itself}. In
principle, this allows the search algorithm, execution schedule, etc. to all be
optimised, replaced or even removed entirely, as long as a proof is found that
it would be beneficial to do so. The \goedelmachine{} concept is framed in the
language of reinforcement learning, with the goal of maximising``reward'' from
some ``utility function'', but it is easily to recast in terms of improving
efficiency: for example, choosing a utility function which provides a one-off
reward of $\frac{1}{t}$ if a result provably-equivalent to the original
program's is returned within $t$ steps; and 0 reward otherwise. The ability to
provide a utility function separate from the initial implementation is also
useful for specifying the desired task more declaratively: for example, if it is
acceptable to deviate from the behaviour of the initial implementation, as long
as some other criteria (given by the utility function) are satisfied. A relevant
example might be altering a search algorithm to find more results faster, even
if those aren't the same results that the original algorithm would have found.

The open-ended search through program and proof space performed by algorithms
like superoptimisation, \hsearch{} and the \goedelmachine{} are similar to that
of Theory Exploration (e.g. the discovery of equivalent expressions by
\isacosy{} and \quickspec{}). One major difference is in \emph{end result} of
these techniques: the latter tends to \emph{build up} structure and abstraction
latent in a set of definitions, whilst the former tend to \emph{tear down} such
layers to find direct, minimalistic implementations.

Another seeming contrast is that Theory Exploration has no specific ``target'',
unlike optimisers: it produces any properties found to match its
``interestingness'' criteria. However, the practical necessity of limiting
superoptimisers to only small snippets of code at a time has resulted in
applications like finding ``peephole optimisations''~\cite{Bansal.Aiken:2006},
whose scattershot nature is more akin to Theory Exploration than traditional
compiler optimisation. Likewise, the improvement in efficiency of an
optimisation (however that is measured or justified) contributes to its
interestingness, and hence suitability as an outputs for Theory Exploration
systems (especially in the use case of a programming assistant).

% TODO: Evolutionary patches
